Towards a Pandemic-Scale Ancestral Recombination Graph of SARS-CoV-2

### 2023-05-10 paper.tex
Updates from Overleaf

\section{Results}
\subsection{Inferred ARGs}
For this preliminary study our goals are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development. Rather
than attempting to be comprehensive we chose to work with a
representative subset of the available data, and limited to
inferences that could be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days
(see Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
represented by the nodes inferred from daily sample clusters
by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
contribute significantly.
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by the it. See Section~\ref{sec:mutation_spectrum}
for more analyis on the patterns of mutations inferred.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
We analyse these events in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For exampl, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
For validation purposes, we compare the genetic
relationships in our ARGs with phylogenetic relationships
inferred by the NextStrain project \citep{Hadfield2018-ef}.
The NextStrain
trees consist of a relatively small set of SARS-CoV-2 genomes, which over the time period
of our ARGs are almost entirely restricted to non-recombinant lineages.
To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
the ARGs to  retain only samples that are in a
specific NextStrain tree. These simplified ``backbone'' ARGs exhibit very
little recombination (6 and 5 trees across the genome for the Wide and Long
ARGs, respectively).
% JK: This leaves a big question hanging - why do we ahave recombination
% in here at all? It's probably just artefactual, right?

@@

Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared
between the wide ARG and a GISAID global ``all-time'' tree from NextStrain
(downloaded on January 21, 2023).
To reduce the ARG to a single tree, we show
the left-most tree (covering the genome interval 0--64) in the Wide ARG;
however, the five other trees in this
simplified ARG show near-identical patterns. Supplementary
Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG.
It is clear that the backbone topology of the sc2ts tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendents form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).

@@

\subsection{Ancestral Recombination Graphs}
\label{sec:args}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination.

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forwards-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines the genomic position at which some variation occurs,
and a \emph{mutation} defines the site, node and derived state
that generates variation.
Given these tables, we can then efficiently construct the local
genealogical trees along the genome (arising from recombination),
and also perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to peformance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to potentially be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
Terasaki2021-geonomics,
Fan2022-genealogical,Korfmann2022-weak,
Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}.

>>>>

\section{Results}
\subsection{Inferred ARGs}
For this preliminary study our goals are to illustrate the utility of
\texttt{sc2ts}, and to closely study the properties of the
inferred ARGs to inform subsequent development. Rather
than attempting to be comprehensive we chose to work with a
representative subset of the available data, limited to
inferences that could be performed on a single server in
a few weeks (see below for further details on timings and
computer hardware used). The cut-off dates for sampling were chosen
arbitrarily.
We inferred two ARGs, which we refer to as the
``Wide'' and ``Long'' ARGs throughout.
The Wide ARG is densely sampled but time-limited and
includes all 1.27 million sample genomes  with a
maximum sampling delay of 30 days
(see Section~\ref{sec:filtering_time_travellers})
up to June 30, 2021.
The Long ARG is more sparsely sampled but
includes data for an additional year (June 30, 2022).
For the long ARG, we use at most 1,000 randomly subsampled
genomes per day from those meeting the maximum sampling delay criterion.

@@

The properties of the inferred ARGs are summarised in Table~\ref{tab:args}.
% TODO more discussion here, making sure we explain anything that might not be
% obvious.
Most of the nodes in the ARGs represent sample genomes
(Wide ARG: 87\%, Long ARG: 84\%), but the inferred ancestral sequences
represented by the nodes inferred from daily sample clusters
by UPGMA (Section~\ref{sec:sample-cluster-tree-inference})
and parsimony heuristics (Section~\ref{sec:parsimony-heuristics})
contribute significantly.
Table~\ref{tab:args} shows the counts of edges (see
Section~\ref{sec:args} for a technical definition) in the ARGs
categorised by parent node type, and demonstrates that each
of these contributes significantly to the overall topology.
In particular, we can see that approximately 40\% of the edges
in both ARGs involve inheriting directly from sampled genomes.
Both ARGs contain 29,422 sites (\ref{sec:data_preprocessing}),
and a large number of mutations.
The average number of mutations per-site is high, although much
of this may be explained by outlier sites with artefactually high
mutation counts (see Figure~\ref{fig:breakpoint-distribution}).
Despite this, however, the number of mutations per node
is small. In the Wide ARG, for example, we have a mean of 0.77
mutations per sample, demonstrating that new samples are generally
well represented by the it. See Section~\ref{sec:mutation_spectrum}
for more analysis of the patterns of inferred mutations.
Recombination plays a relatively minor role, with $<0.3\%$
of nodes in the ARGs representing inferred recombination events.
We analyse these events in Sections~\ref{sec:jackson_recombs},
\ref{sec:breakpoint_intervals}, \ref{sec:parent_divergence},
and \ref{sec:pango_x_lineages}.

Table~\ref{tab:args} also summarises some of the computational
properties of the inferred ARGs.
The ARGs are encoded as a ``succinct tree sequence'' using
the \texttt{tskit} library, which provides an extensive
suite of operations for constructing and analysing ARGs
(Section~\ref{sec:args}). For example, the Wide ARG
which contains complete genomes (with imputed missing data)
for around 1.2 million samples, along with extensive sample
and debugging metadata, requires only 58MB of space (compressed
using the \texttt{tszip} utility). The majority of this
space is used by metadata, which when discarded results in
an encoding that requires an average of only 8.29 bytes per
SARS-CoV-2 genome stored.
Loading these ARGs takes less than a second, and they can be interactively
analysed using Jupyter notebooks on a standard laptop. The majority
of the analyses in this preprint can be carried out in seconds
with the \texttt{tskit} Python API, using a few gigabytes of RAM.

@@

\subsection{Backbone phylogeny}
\label{sec:backbone_phylogeny}
For validation purposes, we compare the genetic
relationships in our ARGs with phylogenetic relationships
inferred by the NextStrain project \citep{Hadfield2018-ef}.
To enable this comparison, we ``simplified''~\citep{Kelleher2018-xc}
the sc2ts ARGs to a ``backbone'' containing only samples present
in a specific NextStrain tree. For this purpose, we picked an
``all-data''  NextStrain tree restricted to the time period covered
by the appropriate ARG. This results in a relatively small
set of shared samples, almost entirely restricted to non-recombinant lineages.
These simplified backbone ARGs exhibit very
little recombination (6 and 5 trees across the genome for the Wide and Long
ARGs, respectively).
% JK: This leaves a big question hanging - why do we ahave recombination
% in here at all? It's probably just artefactual, right?

@@

Figure~\ref{fig:cophylogeny} shows a co-phylogeny of the 88 samples shared
between the wide ARG and a GISAID global ``all-time'' tree from NextStrain
(downloaded on January 21, 2023).
To reduce the ARG to a single tree, we show
the tree in the middle of the viral genome (covering the genome interval 8393 to 21618)
in the Wide ARG;
however, the five other trees in this
simplified ARG show near-identical patterns. Supplementary
Figure~\ref{fig:cophylogeny_long} shows the same comparison for the Long ARG.
It is clear that the backbone topology of the sc2ts tree shows very close
agreement with the NextStrain tree. The sample genomes cluster by their
assigned Pango lineage status, and many variants and their descendents form
identical monophyletic clades in both the trees (e.g., the Alpha and Delta VoC
clades, labelled).

@@

\subsection{Ancestral Recombination Graphs}
\label{sec:args}
The term ``Ancestral Recombination Graph'' was introduced by
Griffiths and colleagues~\citep{Griffiths1991-two,Griffiths1998-ancestral}
and originally defined as an alternative formulation of the coalescent
with recombination stochastic process~\citep{Hudson1983-properties}.
Subsequently, the term ARG came to be used in a more general way to
describe not just realisations of this model, but to any
recombinant genetic ancestry~\citep{Minichiello2006-mapping}.
While there is some subtlety in the details~\citep{Wong2023-efficient},
we can think of an ARG as being any graph that encodes the
reticulate genetic ancestry of a sample of sequences under
the influence of recombination. This definition encompasses various types of graphs traditionally described using the broader term of phylogenetic networks.

The ``succinct tree sequence'' is an ARG data structure
that is both general (in terms of the types of ancestry that can
be described) and computationally efficient~\citep{Wong2023-efficient}.
Originally developed to facilitate large-scale coalescent
simulations~\citep{Kelleher2016-wk}, the methods have been
extended and applied to forwards-time
simulations~\citep{Kelleher2018-xc,Haller2018-tree},
calculation of population genetics statistics~\citep{Ralph2020-efficiently}
and ARG inference~\citep{Kelleher2019-ba,Wohns2022-th}.
The succinct tree sequence is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations. A \emph{node}
represents a particular genome, which may be an observed sample
or an inferred genetic ancestor. The genetic inheritance between
a pair of nodes along a segment of genome is defined by
the \emph{edge} $(\ell, r, p, c)$, which states that
child node $c$ inherited its genome from parent node $p$
from left coordinate $\ell$ to right coordinate $r$. A \emph{site}
defines the genomic position at which some variation occurs,
and a \emph{mutation} defines the site, node and derived state
that generates variation (note that this required that the sequences can be aligned and phased).
Given these tables, we can then efficiently construct the local
genealogical trees along the genome (arising from recombination),
and also perform a range of calculations efficiently by
reasoning about the differences between these local
trees~\citep{Kelleher2016-wk,Ralph2020-efficiently}. These
algorithms have led to performance increases of several orders
of magnitude over the state-of-the-art in a range of
applications~\citep{Kelleher2016-wk,Kelleher2018-xc,Kelleher2019-ba,
Ralph2020-efficiently,Baumdicker2022-ep}.
The succinct tree sequence encoding
is also very concise, allowing, for example, for millions of
complete human genomes to potentially be stored in a few gigabytes of
space~\citep{Kelleher2019-ba}.

The \texttt{tskit} library~\citep{Tskit2023-tskit} is a liberally
licensed open source toolkit that provides a comprehensive suite
of tools for working with ARGs. Based on core functionality written
in C, it provides interfaces in C, Python and Rust. The Python interface
is based on NumPy~\citep{Harris2020-array}, and provides a convenient
platform for interactive analysis of large-scale data using, for
example, Jupyter notebooks~\citep{Kluyver2016-jupyter} and taking
advantage of the analysis tools in the burgeoning PyData ecosystem.
Tskit is mature software, widely used in population genetics, and
has been incorporated into several downstream
applications~\citep[e.g.][]{Haller2019-slim,Speidel2019-yh,
Terasaki2021-geonomics,
Fan2022-genealogical,Korfmann2022-weak,
Mahmoudi2022-bayesian,Petr2022-slendr,Rasmussen2022-espalier}. We emphasise that this ecosystem for storing and manipulating ARGs can generally be used to efficiently record and analyse SARS-CoV-2 genealogies reconstructed using other methods, not only the new approach that we describe here.
